Human capital mediates natural selection in contemporary humans

### 2022-01-03 why-natural-selection.Rmd
Add proof that dN^*/da gets larger at higher levels of human capital

We then show that a simple model of human capital, education and fertility
choices can give rise to these empirical results. To test the theory, we show
that polygenic scores' correlation with fertility is associated with their
correlation with education and earnings. Furthermore, a mediation analysis shows
that part of the correlation with fertility is indeed *via* education. Thus, contemporary natural selection on polygenic scores can be explained by their correlation with earnings-increasing human capital.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, our effect sizes appear unchanged over a large
span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period)), during which government spending on
child-related benefits varied considerably [@hoc1999socsec]. In general, there
is only weak evidence that welfare benefits affect fertility
[@gauthier2007impact]. A second alternative theory is that polygenic scores
correlate with the motivation to have children, i.e. parameter $a$ in the model
[cf. @jones2008fertility]. This theory would not explain why selection effects
are smaller at higher incomes and education levels.[^note] A third alternative
is that traits under selection are linked to externalizing behaviour and
risk-seeking. This might be partially captured by our risk-aversion parameter
$\sigma$, which affects fertility by changing the marginal utility of income,
but a more direct channel is risky sexual behaviour [@mills2021identification].
The data here provide some support for this: scores which might plausibly be
linked to externalizing behaviour, like ADHD and younger age at smoking
initiation, are selected for. However, this theory is less good at explaining
variation in selection across the full range of scores, including physical
measures such as waist-hip ratio and BMI. We test this theory directly by
re-estimating equation \@ref(eq:regression) controlling for a measure of risk
attitude (UK Biobank field 2040).  The median ratio of effect sizes between
regressions with and without controls is `r median(ratio)`; all scores which are
significant at $p < 0.05/33$ in uncontrolled regressions remain so when
controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.
[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$, this
effect becomes stronger, not weaker, at higher wages.

@@

From the meta-analysis of twin studies in @branigan2013variation, we take the
heritability of educational attainment to be around 0.4. This is the variance of
educational attainment explained by the "true" polygenic score for educational
attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
attainment from genetic data. We write the measured polygenic
score of educational attainment, "EA3" from @lee2018gene, as
\[
p=p^{*}+\eta.
\]

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
= \frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}
\equiv \lambda, \label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.

@@

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ does not differ between the two sets of
weighted regressions.

>>>>

We then show that a simple model of human capital, education and fertility
choices can give rise to these empirical results. To test the theory, we show
that polygenic scores' correlation with fertility is associated with their
correlation with education and earnings. Furthermore, a mediation analysis shows
that part of the correlation with fertility is indeed *via* education. Thus, contemporary natural selection on polygenic scores can be explained by scores' correlation with earnings-increasing human capital.

@@

We consider three alternative theories that might explain our results. First,
welfare benefits which incentivize child-bearing might be taken up more among
low-income individuals. However, our effect sizes appear unchanged over a large
span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
\@ref(tab:tbl-children-by-period)), during which government spending on
child-related benefits varied considerably [@hoc1999socsec]. In general, there
is only weak evidence that welfare benefits affect fertility
[@gauthier2007impact]. A second alternative theory is that polygenic scores
correlate with the motivation to have children, i.e. parameter $a$ in the model
[cf. @jones2008fertility]. This theory would not explain why selection effects
are smaller at higher incomes and education levels. (Indeed, in the model, the
effect of $a$ on fertility becomes stronger at higher levels of human capital.) A
third alternative is that traits under selection are linked to externalizing
behaviour and risk-seeking. This might be partially captured by our
risk-aversion parameter $\sigma$, which affects fertility by changing the
marginal utility of income, but a more direct channel is risky sexual behaviour
[@mills2021identification]. The data here provide some support for this: scores
which might plausibly be linked to externalizing behaviour, like ADHD and
younger age at smoking initiation, are selected for. However, this theory is
less good at explaining variation in selection across the full range of scores,
including physical measures such as waist-hip ratio and BMI. We test this theory
directly by re-estimating equation \@ref(eq:regression) controlling for a
measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
sizes between regressions with and without controls is `r median(ratio)`; all
scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
so when controlling for risk attitude. This non-result could simply reflect the
imprecision of the risk attitude measure, which is a single yes/no question.
However, this measure *does* predict the overall number of children, highly
significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
Given that, and the statistical power we get from our sample size, we believe
that the non-result is real: while risk attitude does predict fertility in the
sample, it is not an important channel for natural selection.

@@

From the meta-analysis of twin studies in @branigan2013variation, we take the
heritability of educational attainment to be around 0.4. This is the variance of
educational attainment explained by the "true" polygenic score for educational
attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
attainment from genetic data. We write the measured polygenic
score of educational attainment, "EA3" from @lee2018gene, as
\[
p = p^* + \eta.
\]

For realized educational attainment $EA$, write:
\[
EA = \alpha+\beta p^* + \varepsilon = \alpha + \beta(p - \eta) + \varepsilon
\]
We assume the classical errors-in-variables model, i.e. that $\eta$
is independent of both $p$ and $\varepsilon$. Under these conditions,
the ratio between the estimated $R^{2}$ in a regression of $p$ on
$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
\begin{equation}
\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
= \frac{\sigma_{p^*}^{2}}{\sigma_{p}^{2}}
\equiv \lambda, \label{eq:r2-ratio}
\end{equation}
as we show below. The right hand side is the "reliability ratio".
We regress educational attainment on EA3 in our sample, weighting
by our Age/Qualification weights, to give 
$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.

@@

The errors-in-variables conditions are quite restrictive. In particular, they
rule out that the unmeasured part of PSEA ($\eta$) correlates with the error 
term $\varepsilon$, e.g. with environmental causes of educational attainment. 
If unmeasured PSEA correlates positively with $\varepsilon$, then the
reliability coefficient $\lambda$ will be lower than calculated. However, the
proportional effect of natural selection on the PSEA-income relationship will
remain the same, so long as $\lambda$ does not differ between the two sets of
weighted regressions.
